---
title: "DATA VISUALIZATION(R) PROJECT"
author: "MATRICULATION NUMBER: 22302020 "
date: "2024-06-27"
output: 
  html_document: 
    toc: true
    number_sections: true
editor_options: 
  chunk_output_type: inline
---



# Loading of Libraries.
**All the packages were already installed during the class excercises .
The following libraries were loaded**
```{r}
suppressPackageStartupMessages(library(tidyverse))
library(RColorBrewer)  
library(patchwork)
library(ggrepel)
```



# TCGA expression data:


## Load and prepare the data
```{r}
# Read in the files using the functions discussed in class
#getwd()
setwd("~/Downloads/DataVis_R_2024_FinalProject/")
tcga_cancer <- read_csv("tcga_cancer.csv", show_col_types = FALSE)
tcga_cancer_labels  <- read_csv("tcga_cancer_labels.csv", show_col_types = FALSE)

# select the first 200 samples for gene 1 
tcga_cancer_selected <- tcga_cancer[1:200, 3:45]

# Select the first 200 cancer labels 
tcga_cancer_labels_selected <- tcga_cancer_labels[1:200, ]

# Define data to be used for the ggplot2 by selecting gene_1 by indexing
gene_1 <- tcga_cancer_selected[ , 1]

```

## Plotting:


### Expression distribution Gene1:
a. **Histogram with 0.4 binwidth** 
```{r}
# Define plot 'i' using ggplot, using 'gene_1' data frame for x-axis
i <- ggplot(data.frame(gene_1), aes(x = gene_1)) +
  # Add histogram layer with binwidth of 0.4, black border, and gray fill
  geom_histogram(binwidth = 0.4, color = "black", fill = "gray88") +
  # Set plot title and axis labels
  labs(title = "Expression Gene 1", x = "RNA counts", y = "Count") +
  # Set x-axis and y-axis limits
  coord_cartesian(xlim = c(-0.5, 7), ylim = c(0, 45), expand = FALSE) +
  # Use minimal theme for the plot appearance
  theme_minimal() +
  # Customize plot theme settings
  theme(
    plot.title = element_text(size = 16),  # Title font size
    panel.border = element_rect(fill = NA, linewidth = 2),  # Panel border
    panel.grid = element_blank(),  # Remove panel grid lines
    axis.text = element_text(size = 16, face = "bold"),  # Axis text font size and style
    axis.title = element_text(size = 16, face = "bold"),  # Axis title font size and style
    axis.ticks = element_line(linewidth = 1.5),  # Axis tick marks thickness
    axis.ticks.length = unit(0.1, "cm")  # Axis tick mark length
  )
```

```{r i, fig.width = 4.5, fig.height = 4.55, echo=FALSE, message = FALSE}
plot(i)
```



### Binwidth adjustment
***In histograms, the choice of bin width plays a crucial role in how the data distribution is visualized. A too narrow or too wide bin width can obscure important patterns or details in the data. Typically, the bin width should be chosen such that it effectively represents the underlying distribution without oversmoothing or undersmoothing the data.***
***The above  histogram with 0.4 binwidth setting is moderately smooth, with bins that provide a general overview of the data distribution. The choice of 0.4 seems reasonable for an initial exploration but might benefit from exploring other values to see if a clearer pattern emerges.***




### The histogram with hgher binwidth** 
**The wider bins tend to smooth out the distribution, potentially masking finer details or variability in the data, however the  histogram did not capture subtle peaks or modes as effectively as narrower bin widths.(for example after gene count of 6), leading to loss of important details***
```{r}
# Plot the histogram with 0.5 binwidth.
higher_bin <- ggplot(data.frame(gene_1), aes(x = gene_1)) +
  geom_histogram(binwidth = 0.5, color = "black", fill = "lightgray" ) +
  labs(title = "Expression Gene 1", x = "RNA counts", y = "Count") +
  coord_cartesian(xlim = c(-0.5, 7), ylim=c(0, 45), expand = FALSE) +
  theme_minimal()+
  theme(
    plot.title = element_text(size = 16),
    panel.border = element_rect(fill = NA, linewidth = 2),
    panel.grid = element_blank(), 
    axis.text = element_text(size = 16, face = "bold"),  
    axis.title = element_text(size = 16, face ="bold"),  
    axis.ticks  = element_line(linewidth  = 1.5),  
    axis.ticks.length = unit(0.1, "cm"), 
  )
```

```{r higher_bin, fig.width = 4.5, fig.height = 4.55, echo=FALSE, message = FALSE}
plot(higher_bin)
```




### The histogram with lower binwidth
***With a bin width of 0.2, the histogram provides more bins, potentially revealing finer details in the distribution of gene expression counts. This finer granularity might be beneficial for detecting subtle peaks or modes that were less apparent with the wider bin width of 0.4. However, it can also introduce more noise or variability.***
```{r}
lower_bin <- ggplot(data.frame(gene_1), aes(x = gene_1)) +
  geom_histogram(binwidth = 0.2, color = "black", fill = "lightgray" ) +
  labs(title = "Expression Gene 1", x = "RNA counts", y = "Count") +
  coord_cartesian(xlim = c(-0.5, 7), ylim=c(0, 45), expand = FALSE) +
  theme_minimal()+
  theme(
    plot.title = element_text(size = 16),
    panel.border = element_rect(fill = NA, linewidth = 2),
    panel.grid = element_blank(), 
    axis.text = element_text(size = 16, face = "bold"), 
    axis.title = element_text(size = 16, face ="bold"),  
    axis.ticks  = element_line(linewidth  = 1.5), 
    axis.ticks.length = unit(0.1, "cm"),
  )
```

```{r lower_bin , fig.width = 4.5, fig.height = 4.55, echo=FALSE, message = FALSE}
plot(lower_bin)
```



### Gene expression correlation
```{r}
# Define part of data to be used for the plot by selecting Gene_12 by indexing(Gen-1 has been selected above)
gene_12 <- tcga_cancer_selected[ , 12]

# Pearson correlation coefficient estimation
correlation <- cor(gene_1, gene_12) 

# Scatter plot with linear regression line and Pearson's correlation:
ii <- ggplot(tcga_cancer_selected, aes(x = gene_1, y = gene_12)) +
  
  # Add linear regression line with dashed line type, black color, and standard error shading
  
  geom_smooth(method = "lm", color = "black", linetype = "dashed", se = TRUE) +
  
  # Add points with solid black border, light cyan fill, size 4, and 0.5 stroke width
  geom_point(shape = 21, color = "black", fill = "#99cccc", size = 4, stroke = 0.5) +
  
  # Annotate Pearson's correlation coefficient with rounded to 3 decimals
  annotate("text", x = Inf, y = Inf, label = paste("R = ", round(correlation, 3)), hjust = 1.1, vjust = 3, size = 5) +
  
  # Set plot title, x-axis label, and y-axis label
  labs(
    title = "Relationship Gene 1 and Gene 12",
    x = "Gene 1",
    y = "Gene 12"
  ) +
  
  # Use theme_bw() for a black and white theme
  theme_bw() +
  
  # Customize theme elements
  theme(
    plot.title = element_text(size = 16),  # Title font size
    panel.border = element_rect(linewidth = 3),  # Panel border thickness
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.ticks = element_line(linewidth = 1),  # Axis tick marks thickness
    axis.text = element_text(size = 16),  # Axis text size
    axis.title = element_text(size = 16, face = "bold")  # Axis title size and bold face
   
        )
```

```{r ii, fig.width = 4.5, fig.height = 4.55, echo=FALSE, message = FALSE}
plot(ii)
```
 


### Class proportions (Stacked Bar Chart)
```{r}
#Select the first 200 cancer labels and have a glimpse of it
tcga_cancer_labels_selected <- tcga_cancer_labels[1:200, ]

# Combine data with labels
combined_data <- cbind(tcga_cancer_labels_selected, tcga_cancer_selected)

# Convert the labels to factors
combined_data$Class <- as.factor(combined_data$Class)

# Summarize the data to get counts of each class
class_counts <- combined_data %>%
  group_by(Class) %>%
  summarise(Count = n())

# Calculate percentages
class_counts <- class_counts %>%
  mutate(Percentage = Count / sum(Count) * 100)

# Define  aquantive color palette("Pastel1" ) for the 3 classes and flipping the color for the classes
flipped_colors <- rev(brewer.pal(n = length(unique(class_counts$Class)), name = "Pastel1"))

# Create a stacked bar plot showing class distribution
iii <- ggplot(class_counts, aes(x = Percentage, y = "", fill = Class)) + 
  # Define data and aesthetics
  geom_bar(stat = "identity", width = 0.5, color = "black") + 
  # Create a bar plot with black borders
  coord_cartesian(expand = FALSE) +  
  
  # Ensure plot limits match data without expansion
  geom_text(aes(label = Count),  # Add text labels for bar counts
            position = position_stack(vjust = 0.5),  
            hjust = 0.5,  # Center labels horizontally
            size = 5, color = "black") +  # Set label size and color
  scale_fill_manual(values = flipped_colors) +  # Use custom fill colors
  labs(
    x = "Percentage (%)",  # X-axis label
    y = NULL,  # No y-axis label
    title = "Distribution of classes (n = 200)"  # Plot title
  ) +
  theme_minimal() +  # Use minimal theme
  theme(
    axis.text = element_text(size = 16, face = "bold"),  # Axis text size and bold face
    
    axis.title = element_text(size = 16, face ="bold"),  # Axis title size and bold face
    
    plot.title = element_text(size = 16),  # Plot title size
    axis.ticks.x  = element_line(linewidth  = 1.5),  # X-axis tick thickness
    axis.ticks.length = unit(0.1, "cm"),  # X-axis tick length
    legend.position = "top",  # Position legend at the top
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    aspect.ratio = 0.12  # Set aspect ratio
  ) +
  guides(fill = guide_legend(reverse = TRUE))  # Reverse the legend order

iii  
```





### Correlation:
**Correlation is an estimation of the strength(magnitude) of the linear relationship between two variables. In correlation analysis. The Pearson correlation coefficient(which ranges from -1 to 1)  with values toward 1 being interpreted as strong positive linear relationships  , values toward -1 as  negative but strong linear relationship, and  0 indicates no linear relationship.**
**Calculated coefficient suggest very weak negative linear correlation (as one variable increases, the other tends to decrease slightly) between Gene1 and Gene12.**
  


### Combine the plots for plot i, ii and iii
```{r}
# Combine plots i and ii horizontally with specified relative widths
horizontal_top <- i + ii + plot_layout(widths = c(1, 2))

# Combine the horizontal_top with plot iii stacked vertically
final_plot <- horizontal_top / iii + plot_layout(heights = c(2, 1)) + 
 
# Annotate the combined plot in order to set the numbering i, ii and iii respectively
 plot_annotation(tag_levels = list(c("i", "ii", "iii")))
```


```{r final_plot , fig.width = 8, fig.height = 8, echo=FALSE, message = FALSE}
plot(final_plot)
```





# Human genome :



## Load and prepare data
```{r}
# Read in the files using the functions discussed in class
getwd()
setwd("~/Downloads/DataVis_R_2024_FinalProject/")
genes <- read.delim("hg38_genes.tsv", header = TRUE, stringsAsFactors = FALSE)

# Filter for specific gene types
filtered_gene_types <- genes %>%
  filter(gene_type %in% c("protein_coding", "lncRNA", "processed_pseudogene", "transcribed_unprocessed_pseudogene"))

# Rename "protein_coding" to "mRNA" and combine pseudogene types
filtered_renamed_gene <- filtered_gene_types %>% 
mutate(gene_type = recode(gene_type, 
                          "protein_coding" = "mRNA", 
                          "processed_pseudogene" = "pseudo", 
                          "transcribed_unprocessed_pseudogene" = "pseudo"))
# Estimation of gene_lenth
gene_start <- filtered_renamed_gene[, 2]
gene_end <- filtered_renamed_gene[, 3]
gene_length2 <- gene_end - gene_start

# Filter by gene length
filtered_renamed_gene <- filtered_renamed_gene %>%
  filter( gene_length2 > 1000 & gene_length2 <= 20000)

# Summarize counts per gene class
gene_counts <- filtered_renamed_gene %>%
  group_by(gene_type) %>%
  summarise(count = n())

# Reverse the order of the classes
gene_counts$gene_type <- factor(gene_counts$gene_type, levels = rev(unique(gene_counts$gene_type)))

# Define the Set2 color palette and reverse it
set2_colors <- rev(brewer.pal(n = 3, name = "Set2"))
```




## Plotting.


### Gene count(Bar Chart).
```{r}
# Importing the ggplot2 library and setting up the plot
A <- ggplot(gene_counts, aes(x = gene_type, y = count, fill = gene_type)) +
  
  # Creating a bar plot where the heights of the bars represent counts
  geom_bar(stat = "identity") +
  
  # Setting the y-axis limit to a maximum slightly higher than the highest count
  ylim(0, max(gene_counts$count) * 1.1) +
  
  # Setting the Cartesian coordinate limits for both x and y axes, removing padding
  coord_cartesian(xlim = c(0.5, 3.5), ylim=c(0, 4000), expand = FALSE) +
  
  # Manually setting the fill colors for the bars
  scale_fill_manual(values = set2_colors) +
  
  # Manually setting the colors of the outlines of the bars
  scale_color_manual(values = set2_colors) +
  
  # Adding labels and a title to the plot
  labs(
    title = "Total gene number",
    x = "Class",
    y = "Count"
  ) +
  
  # Applying a minimal theme to the plot for a cleaner look
  theme_minimal() +
  
  # Customizing various theme elements of the plot
  theme(
    # Customizing the plot title
    plot.title = element_text(size = 16, face = "bold"),
    
    # Adding dashed lines to the major y-axis grid lines
    panel.grid.major.y = element_line(size = 0.5, linetype = "dashed", colour = "grey"),
    
    # Removing minor grid lines on the x-axis
    panel.grid.minor.x = element_blank(),
    
    # Removing major grid lines on the x-axis
    panel.grid.major.x = element_blank(),
    
    # Removing minor grid lines on the y-axis
    panel.grid.minor.y = element_blank(),
    
    # Adding a border around the panel
    panel.border = element_rect(fill = NA, linewidth = 0.5),
    
    # Rotating and customizing the x-axis text for better readability
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    
    # Customizing the y-axis text
    axis.text.y = element_text(angle = 0, size = 14),
    
    # Customizing the axis titles
    axis.title = element_text(size = 16, face = "bold"),
    
    # Removing the legend as it is not needed
    legend.position = "none"
  )

```

```{r A, fig.width = 4.5, fig.height = 4.5, echo=FALSE, message = FALSE}
plot(A)
```



### Length distribution.
```{r}
#To obtain gene_lenght 
gene_start <- filtered_renamed_gene[, 2]
gene_end <- filtered_renamed_gene[, 3]
gene_length2 <- gene_end - gene_start

# Calculate medians for each gene type
medians <- aggregate(gene_length2 ~ gene_type, filtered_renamed_gene, median)

# Define the color for gene categories 
set_color  <- c("pseudo" ="#8DA0CB", "mRNA" = "#FC8D62", "lncRNA" = "#66C2A5")

# Plot the length density distribution of the gene classes
B <- ggplot(filtered_renamed_gene, aes(x = gene_length2, fill = gene_type)) + # Define data and aesthetics
  geom_density(alpha = 0.5) +  # Add density plot with transparency
  geom_vline(data = medians, aes(xintercept = gene_length2, color = gene_type), linetype = "dashed") +  # Add dashed vertical lines for medians
  labs(x = "Gene Length(bp)", y = "Density", title = "Length distribution of gene classes") +  # Set labels and title
  scale_fill_manual(values = set_color) +  # Manually set fill colors
  scale_color_manual(values = set_color) +  # Manually set line colors
  scale_y_continuous(expand = c(0, 0)) +  # Ensure no expansion of y-axis limits
  theme_minimal() +  # Apply minimal theme
  theme(
    plot.title = element_text(size = 16),  # Set plot title size
    axis.title = element_text(face = "bold", size = 16),  # Set axis title font style and size
    axis.text = element_text(size = 16, hjust = 1),  # Set axis text size and horizontal justification
    panel.grid.minor = element_blank()  # Remove minor grid lines
  )
```

```{r B, fig.width = 8, fig.height = 4.5, echo=FALSE, message = FALSE}
plot(B)
```

**The overlapping areas  in the above plot type is making it difficult to distinguish between classes, hence it Less effective for comparing exact quartiles and outliers.**

*The box plot is generally more intuitive to understand and read, especially for comparing central tendencies, variability, and identifying outliers across classes. It provides clear visual markers for median, quartiles, and outliers without the potential confusion caused by overlapping densities.**



**Box plot as an alternative plot for length distribution of the gene classes**
```{r}
# Define the box plot
box_plot <- ggplot(filtered_renamed_gene, aes(x = gene_type, y = gene_length2, fill = gene_type)) +  # Define data and aesthetics
  geom_boxplot() +  # Add box plot
  labs(x = "Gene Classes", y = "Gene Length", title = "Box Plot of Gene Length by Class") +  # Set labels and title
  scale_fill_manual(values = c("pseudo" ="#8DA0CB", "mRNA" = "#FC8D62", "lncRNA" = "#66C2A5")) +  # Manually set fill colors
  scale_color_manual(values = c("pseudo" ="#8DA0CB", "mRNA" = "#FC8D62", "lncRNA" = "#66C2A5")) +  # Manually set line colors
  theme_minimal() +  # Apply minimal theme
  theme(
    plot.title = element_text(size = 16),  # Set plot title size
    axis.title = element_text(face = "bold", size = 16),  # Set axis title font style and size
    axis.line = element_line(),  # Set axis lines
    axis.text = element_text(size = 16, hjust = 1, face = "bold"),  # Set axis text size, horizontal justification, and font style
    panel.border = element_rect(fill = NA, linewidth = 0.5),  # Set panel border
    panel.grid.minor.y = element_blank()  # Remove minor grid lines on y-axis
  )
```

```{r box_plot, fig.width = 6, fig.height = 6, echo=FALSE, message = FALSE}
plot(box_plot)
```




### Included figure(combined Plot).
```{r}
# Combine plots A and B horizontally with specified relative widths
combined_plot <- A + B + plot_layout(widths = c(1, 2))

# Annotate the combined plot to set the numbering A and B respectively
final_plot2 <- combined_plot + plot_annotation(tag_levels = list(c("A", "B")))

```

```{r final_plot2, fig.width = 12, fig.height = 6, echo=FALSE, message = FALSE}
plot(final_plot2)
```



# Differential gene expression:



## 5.1 Load and prepare data.
```{r}
# Change working directory.
setwd("~/Downloads/DataVis_R_2024_FinalProject/")

# Read the .txt file.
DEseq2_data <- read.table("deseq2_dummy.txt", header = TRUE, sep = "\t")

# Calculate the average of normalized counts for each gene across all samples (normal and cancer).
DEseq2_data_with_av_normalized_count <- DEseq2_data %>%
  rowwise() %>%
  mutate(avg_normalized_count = mean(c_across(9:14), na.rm = TRUE)) %>%
  ungroup()

# Select the top 2000 genes based on the average normalized count values.
top_2000_genes <- DEseq2_data_with_av_normalized_count %>%
  arrange(desc(avg_normalized_count)) %>%
  slice_head(n = 2000)

# Display the first few rows of the top 2000 genes
view(top_2000_genes)

# Add a column to indicate significant genes based on adjusted p-value and log2 fold change
top_2000_genes <- top_2000_genes %>%
  mutate(significant = case_when(log2FoldChange > 0 & padj < 0.1 ~ "up",
                                 log2FoldChange < 0 & padj < 0.1 ~ "down",
                                 TRUE ~ "n.s"))

# Add a column to label significant genes with log2 fold change higher than 2
top_2000_genes_label <- top_2000_genes %>%
  mutate(label = ifelse(padj < 0.1 & log2FoldChange > 2, gene_id, NA))

```



## Plotting



### Volcano plots
```{r}
# Create the volcano plot
volcano_plot <- ggplot(top_2000_genes, aes(x = log2FoldChange, y = -log10(padj))) +  # Define data and aesthetics
  geom_point(aes(fill = significant), shape = 21, size = 2) +  # Add points with fill based on 'significant' column
  scale_fill_manual(values = c("up" = "brown", "down" = "skyblue", "n.s" = "white")) +  # Manually set fill colors
  labs(title = "Diff. expressed genes (Cancer vs. Normal)",  # Set plot title and axis labels
       x = "Log2 Fold Change",
       y = "Adjusted p-value (-log10)",
       fill = "Regulation") +  # Set fill legend title
  theme_minimal() +  # Apply minimal theme
  geom_label_repel(aes(label = top_2000_genes_label$label),  # Add labels with repulsion to avoid overlapping
                   nudge_y = 0.5,  # Adjust label position vertically
                   nudge_x = 0.5,  # Adjust label position horizontally
                   box.padding = 0.5,  # Set padding around labels
                   point.padding = 0.5,  # Set padding around points
                   na.rm = TRUE,   # Remove missing values
                   max.overlaps = Inf) +  # Set maximum overlaps for labels
  geom_vline(xintercept = 0, linewidth = 0.5) +  # Add vertical line at x = 0
  geom_hline(yintercept = 1, linetype = "dotted", linewidth = 0.5) +  # Add horizontal line at y = 1 with dotted line
  coord_cartesian(xlim = c(-2, 5), ylim=c(0.4, 10), expand = TRUE) +  # Set limits for x and y axes
  theme(
    plot.title = element_text(size = 16),  # Set plot title size
    axis.title = element_text(face = "bold", size = 16),  # Set axis title font style and size
    axis.text = element_text(size = 16, face = "bold", hjust = 1),  # Set axis text size, font style, and horizontal justification
    panel.grid.major.y = element_blank(),  # Remove major grid lines on y-axis
    panel.grid.minor.x = element_blank(),  # Remove minor grid lines on x-axis
    panel.grid.major.x = element_blank(),  # Remove major grid lines on x-axis
    panel.grid.minor.y = element_blank(),  # Remove minor grid lines on y-axis
    panel.border = element_rect(linewidth = 2, fill = NA),  # Set panel border
    legend.position = c(0.91, 0.91),  # Set legend position
    legend.box.background = element_rect(color = "black", fill = NA, linewidth = 1.5),  # Set legend box background
    legend.key.width = unit(1.2, "cm")  # Set legend key width
  )

```

```{r volcano_plot, fig.width, fig.height = 8.5, echo=FALSE, message = FALSE}
plot(volcano_plot)
````



### Heatmaps
```{r}
deseq2_res_f <- top_2000_genes %>% #change deseq2_dummy to  subset above
 
slice_min(n = 20, order_by = padj) #i changed "padj" to padj without the quotes

# Reshape the data for plotting
deseq2_res_long <- deseq2_res_f %>%
  dplyr::select(gene_id, 9:14) %>%
  pivot_longer(cols = 2:7, names_to = "condition", values_to = "counts") #changed cals to 2:7 instead of 9:14

# Create the heatmap
Heatmaps <- ggplot(deseq2_res_long, aes(x = condition, y = gene_id, fill = log2(counts))) +  # Define data and aesthetics
  geom_tile(color = "black") +  # Add tiles with black borders
  geom_text(aes(label = round(log2(counts), digits = 2))) +  # Add text labels with rounded log2 counts
  coord_cartesian(expand = FALSE) +  # Set Cartesian coordinates with no expansion
  scale_fill_distiller(palette = "Purples") +  # Set color palette for fill
  labs(x = "Sample", y = "Gene IDs", fill = "Counts (log2)") +  # Set axis and fill legend labels
  theme_minimal() +  # Apply minimal theme
  theme(
    legend.position = "top",  # Set legend position to top
    legend.ticks = element_blank(),  # Remove legend ticks
    title = element_text(size = 14),  # Set title text size
    axis.text = element_text(size = 11, colour = "black"),  # Set axis text size and color
    axis.title = element_text(size = 15, face = "bold"),  # Set axis title size and font
    panel.border = element_rect(linewidth = 3, fill = NA)  # Set panel border properties
  )

````

```{r Heatmaps, fig.width, fig.height = 8.5, echo=FALSE, message = FALSE}
plot(Heatmaps)
````

